Numerical modelling of the Cordilleran ice sheet

### 2014-08-06 kappa.tex
Regroup all parameter tables into one.

@@

This section does not attempt to provide an extensive description of PISM,
which can be found in an on-line documentation
\citep[\url{http://www.pism-docs.org},][]{PISM-authors.2014}.
Instead, it aims to highlight the modelling choices that I have made in this
thesis, and to present an overview of the physics involved in these choices.

@@

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
and the deviatoric stress tensor, ${\DST = \CST + p\,\IDT}$,
where $p=-\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$, and
water content, $\omega$, through an Arrhenius-type law
\citep[Eqs.~63--65]{Paterson.Budd.1982, Aschwanden.etal.2012},
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. $R$ is the ideal gas constant,
and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameter corresponding to
values measured below and above a critical temperature threshold
${T_c=-10}$\,\degC\ \citep[\cref{tab:iceflowparams};][]{Paterson.Budd.1982}.
The water fraction, $\omega$, is capped at a maximum value of 0.01, above which
no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
Greve.1997}.

Finally, the evolution of temperature within the ice is governed by the
\emph{conservation of energy}. In PISM, the conservation of energy uses an
enthalpy formulation in order to
account for thermodynamic effects associated with internal phase changes in the
glacier \citep[Eqs.~20--21]{Aschwanden.etal.2012}. This enthalpy ($H$)
formulation reads
\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\matdv{H}\right)
        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4} \,,
\end{equation}
where $\vect{q}$ represents the heat flow and
${\nu\dot{\epsilon_e}^2/4 = \tr(\DST\SRT)}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, the heat flow is
modelled by Fourier's law, $\vect{q} = k\,\grad{T}$, where $k$ is the thermal
conductivity of ice, while enthalpy can be expressed as $H=c\,T$, where $c$
denotes the ice specific heat capacity (\cref{tab:iceflowparams}).
Consequently, the enthalpy equation (\ref{eqn:enthalpy}) can be rewritten using
a more familiar, cold-ice, temperature formulation,
\begin{equation}
    \label{eqn:temperature}
    \matdv{T} = \frac{k}{\rho c} \Delta T
                + \frac{\nu\doteps_e^2}{4\rho c} \,.
\end{equation}
However, note that, here, the term ``enthalpy'' is actually an abuse of
language, commonly used in glaciology as a synonym of ``interval energy''
\citep{Aschwanden.etal.2012}.

\begin{table*}
  \centering
  \caption{Constant parameters of the ice flow model.}
  \label{tab:iceflowparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

@@

    \hline
  \end{tabular*}}
\end{table*}

@@

The simulations presented here use a \emph{pseudo-plastic} sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{{v_{th}}^q\,|\vv_b|^{1-q}} \,,
\end{equation}
where $\vect{\tau}_b$ is the basal traction force, $\vv_b$ is the basal
velocity, and $v_{th}$ is an arbitrary velocity threshold
(\cref{tab:slidingparams}). Note that, in the case of
$q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law
\citep[Eq.~11]{Winkelmann.etal.2011},
\begin{equation}
    \label{eqn:plastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{|\vv_b|} \,.
\end{equation}

@@

In {\CCYC}, using PISM~0.6, the effective pressure is determined by an
empirical parametrisation based on laboratory experiments on Antarctic till
\citep{Tulaczyk.etal.2000, Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $e_0$ is a measured reference void ratio, $C_c$ a measured
compressibility coefficient, and $\delta$ is a non-dimensional parameter
chosen as 0.02 based on simple sensitivity tests on the Greenland ice sheet
(Table~\ref{tab:slidingparams}).

Finally, the till friction angle, $\phi$ (\cref{eqn:coulomb}), is taken as a
function of modern bed
elevation, with lowest values occurring at low elevations, thus accounting for
a weakening of till associated with the presence of marine sediments
\citep[cf.][Eq.~10]{Martin.etal.2011},
\begin{equation}
    \phi(x,y) =
    \begin{cases}
        \phi_0 & \text{if}\ b \le b_0 \,, \\
        \phi_0 + (\phi_1-\phi_0) \frac{b - b_0}{b_1-b_0}
                & \text{if}\ b_0 < b < b_1 \,, \\
        \phi_1 & \text{if}\ b_1 \le b \,,
    \end{cases}
\end{equation}
where $b_0=0$\,m is modern sea-level and $b_1=200$\,m is a rough average of
highest observed palaeo-shoreline elevations in the model domain
\citep[\cref{tab:slidingparams};][]{Clague.James.2002}.

\begin{table*}
  \centering
  \caption{Constant parameters of the sliding model. Missing sources
           correspond to uncalibrated values.}
  \label{tab:slidingparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

    \hline
  \end{tabular*}}
\end{table*}

@@

The modelled ice sheet interacts with the underlying bedrock in two ways.
Firstly, \emph{bedrock temperatures} are computed to a given depth by
appending the
ice enthalpy model (\ref{eqn:enthalpy}) with an underlying bedrock thermal
model. The only process accounted for is vertical heat conduction, hence this model
consists of a simple diffusion equation,
\begin{equation}
    \pdv{T}{t} = \frac{k_b}{\rho_b c_b} \Delta T \,,
\end{equation}
where $k_b$, $\rho_b$ and $c_b$ are the thermal conductivity, density and
specific heat capacity of the bedrock (\cref{tab:bedrockparams}).
Bedrock temperature is conditioned at depth by a fixed upwards geothermal heat
flux $q_G$  (\cref{tab:bedrockparams}). The bedrock thermal model
is necessary for modelling the Cordilleran ice sheet on multi-millennial
time-scales, because temperature fluctuations caused by climate change and
varying degrees of isolation from atmospheric temperatures by ice sheet
coverage, penetrate several kilometres
into the rock, and eventually affect basal ice temperatures.

Secondly, bedrock elevation evolves in response to the ice weight. The
\emph{bedrock deformation} model used in this thesis describes the flexure of
an elastic
lithosphere on top of an infinite half-space viscous mantle
\citep{Lingle.Clark.1985}. It can be described by a single pseudo-differential
equation of the bed elevation, $b$,
\begin{equation}
    \label{eqn:beddef}
    2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb} \,,
\end{equation}
where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity (\cref{tab:bedrockparams}), and $\sigma_{zzb}$
corresponds to the ice load
\citep{Bueler.etal.2007}. Here, ${\Delta^2 = (\div\grad{})^2}$ designs the
biharmonic (Laplacian square) operator of
linear elastic theory, while $|\grad{}|$ is a pseudo-differential operator
defined through its Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. On
the left-hand side of \cref{eqn:beddef}, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response.

\begin{table*}
  \centering
  \caption{Constant parameters of the bedrock thermal and deformation models.}
  \label{tab:bedrockparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

    \hline
  \end{tabular*}}
\end{table*}

@@

In the simulations presented here, \cref{eqn:calovgreve} is numerically
approximated using week-long sub-intervals. For each sub-interval, PDDs, which
serve as a proxy for the available energy for melt, define the amount of snow
melt through a proportionality factor $F_s$. If all snow is melted, remaining
PDDs are then used to melt ice, using a proportionality factor $F_i$
(\cref{fig:model-pdd}). The degree-day factors for snow, $F_s$, and ice, $F_i$
(\cref{tab:surfaceparams}),
are derived from mass-balance measurements on contemporary glaciers in the
Coast and Rocky mountains of British Columbia \citep{Shea.etal.2009}. In
{\CCLI}, the temperature standard deviation $\sPDD$ is a constant model
parameter. In {\CCYC}, it is computed from daily temperature values from the
North American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, using
a method initially developed in {\PSDV}, and further improved upon in {\PSDP}.
Although the PDD model is a crude parametrisation of processes happening at the
surface of a glacier, its applicability has been demonstrated
\citep[e.g.][]{Hock.2003},
while other, physically-based methods \citep[e.g.][]{Hock.2005}, require a
precise knowledge of the physical state of the atmosphere, which typically is highly
uncertain for past periods such as the last glacial cycle.

@@

\begin{table*}
  \centering
  \caption{Constant parameters of the surface and atmosphere models.}
  \label{tab:surfaceparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

    \hline
  \end{tabular*}}
\end{table*}

@@

The present-day monthly climatology is computed from
near-surface air temperature and precipitation rate fields extracted from
observational data and atmospheric reanalyses (\cref{tab:datasources}).
The lapse-rate correction, ${\Delta}T_{LR}$, is computed from the ice sheet
surface elevation, $s$, in relation to the
climate input reference bedrock topography, $b_{ref}$,
\begin{align}
    {\Delta}T_{LR}(t, x, y) = -\gamma\,[&s(t, x, y) - b_{ref}(x, y)] \\
                            = -\gamma\,[&h(t, x, y) + b(t, x, y) \notag \\
                                        &- b_{ref}(x, y)] \,,
\end{align}
thus accounting for the evolution of ice thickness, ${h=s-b}$, on the one hand,
and
for differences between the the ice flow model basal topography, $b$, and the
climate input topography, $b_{ref}$, on the other hand. All simulations use an
annual temperature lapse rate of $\gamma = 6\,\unit{K\,km^{-1}}$
(\cref{tab:surfaceparams}).

\begin{table*}
  \centering
  \caption{Sources of forcing data used to inform the ice sheet model.}
  \label{tab:datasources}
  \makebox[0pt]{\begin{tabular*}{160mm}{@{\extracolsep{\fill}}llll}
    \hline
    Name        & Type & Original reference & Employed in \\
    \hline

@@

    \hline
  \end{tabular*}}
\end{table*}

@@

%\begin{table}
%  \centering
%  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
%           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
%           PROJ.4 parameter names, as used in all model runs, ice volume and
%           area calculations, and most figures.}
%  \label{tab:projparams}
%  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
%    \hline
%    Parameter               & Name              & Value \\
%    \hline
%    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
%    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
%    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
%    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
%    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
%    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
%    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
%    \hline
%  \end{tabular*}}
%\end{table}

@@

Next to the atmospheric forcing, another critical source of uncertainty in
palaeo-ice sheet modelling is the choice of parameters used by the surface mass
balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
are the degree-day factors for snow, $F_s$, and ice, $F_i$
(\cref{fig:model-pdd,tab:surfaceparams}), and the standard deviation of
near-surface air temperature, $\sPDD$
(\cref{eqn:calovgreve,tab:surfaceparams}). Although the first two can only be
deduced from surface mass-balance measurements on present-day glaciers
\citep[e.g.,][]{Shea.etal.2009} and assumptions that these values are also
valid for palaeo-simulations, $\sPDD$ can be estimated from temperature time
series \citep[e.g.,][]{Fausto.etal.2011}.

>>>>

% Copernicus-like vertical space in tables
\newcommand\tophline{\hline\noalign{\vspace{1mm}}}
\newcommand\middlehline{\noalign{\vspace{1mm}}\hline\noalign{\vspace{1mm}}}
\newcommand\bottomhline{\noalign{\vspace{1mm}}\hline}

@@

This section does not attempt to provide an extensive description of PISM,
which can be found in an on-line documentation
\citep[\url{http://www.pism-docs.org},][]{PISM-authors.2014}.
Instead, it aims to highlight the modelling choices that I have made in this
thesis, and to present an overview of the physics involved in these choices.
Parameter values are summarised in \cref{tab:params}.

@@

\begin{table*}
  \centering
  \caption{Constant parameters of the ice sheet model.}
  \label{tab:params}
  \makebox[0pt]{\begin{tabular*}{170mm}{@{\hspace{2em}}l@{\extracolsep{\fill}}lrll}
    \tophline
    Not.    & Name & Value & Unit & Source \\
    \middlehline

    \multicolumn{2}{l}{\emph{Ice material properties}} \\

@@

    \multicolumn{2}{l}{\emph{Basal sliding}} \\

    \multicolumn{2}{l}{\emph{Bedrock and lithosphere}} \\

    \multicolumn{2}{l}{\emph{Surface and atmosphere}} \\

    \bottomhline
  \end{tabular*}}
\end{table*}

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
and the deviatoric stress tensor, ${\DST = \CST + p\,\IDT}$,
where $p=-\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$, and
water content, $\omega$, through an Arrhenius-type law
\citep[Eqs.~63--65]{Paterson.Budd.1982, Aschwanden.etal.2012},
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. $R$ is the ideal gas constant,
and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameter corresponding to
values measured below and above a critical temperature threshold
${T_c=-10}$\,\degC\ \citep[\cref{tab:params};][]{Paterson.Budd.1982}.
The water fraction, $\omega$, is capped at a maximum value of 0.01, above which
no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
Greve.1997}.

Finally, the evolution of temperature within the ice is governed by the
\emph{conservation of energy}. In PISM, the conservation of energy uses an
enthalpy formulation in order to
account for thermodynamic effects associated with internal phase changes in the
glacier \citep[Eqs.~20--21]{Aschwanden.etal.2012}. This enthalpy ($H$)
formulation reads
\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\matdv{H}\right)
        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4} \,,
\end{equation}
where $\vect{q}$ represents the heat flow and
${\nu\dot{\epsilon_e}^2/4 = \tr(\DST\SRT)}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, the heat flow is
modelled by Fourier's law, $\vect{q} = k\,\grad{T}$, where $k$ is the thermal
conductivity of ice, while enthalpy can be expressed as $H=c\,T$, where $c$
denotes the ice specific heat capacity (\cref{tab:params}).
Consequently, the enthalpy equation (\ref{eqn:enthalpy}) can be rewritten using
a more familiar, cold-ice, temperature formulation,
\begin{equation}
    \label{eqn:temperature}
    \matdv{T} = \frac{k}{\rho c} \Delta T
                + \frac{\nu\doteps_e^2}{4\rho c} \,.
\end{equation}
However, note that, here, the term ``enthalpy'' is actually an abuse of
language, commonly used in glaciology as a synonym of ``interval energy''
\citep{Aschwanden.etal.2012}.

@@

The simulations presented here use a \emph{pseudo-plastic} sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{{v_{th}}^q\,|\vv_b|^{1-q}} \,,
\end{equation}
where $\vect{\tau}_b$ is the basal traction force, $\vv_b$ is the basal
velocity, and $v_{th}$ is an arbitrary velocity threshold
(\cref{tab:params}). Note that, in the case of
$q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law
\citep[Eq.~11]{Winkelmann.etal.2011},
\begin{equation}
    \label{eqn:plastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{|\vv_b|} \,.
\end{equation}

@@

In {\CCYC}, using PISM~0.6, the effective pressure is determined by an
empirical parametrisation based on laboratory experiments on Antarctic till
\citep{Tulaczyk.etal.2000, Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $e_0$ is a measured reference void ratio, $C_c$ a measured
compressibility coefficient, and $\delta$ is a non-dimensional parameter
chosen as 0.02 based on simple sensitivity tests on the Greenland ice sheet
(Table~\ref{tab:params}).

Finally, the till friction angle, $\phi$ (\cref{eqn:coulomb}), is taken as a
function of modern bed
elevation, with lowest values occurring at low elevations, thus accounting for
a weakening of till associated with the presence of marine sediments
\citep[cf.][Eq.~10]{Martin.etal.2011},
\begin{equation}
    \phi(x,y) =
    \begin{cases}
        \phi_0 & \text{if}\ b \le b_0 \,, \\
        \phi_0 + (\phi_1-\phi_0) \frac{b - b_0}{b_1-b_0}
                & \text{if}\ b_0 < b < b_1 \,, \\
        \phi_1 & \text{if}\ b_1 \le b \,,
    \end{cases}
\end{equation}
where $b_0=0$\,m is modern sea-level and $b_1=200$\,m is a rough average of
highest observed palaeo-shoreline elevations in the model domain
\citep[\cref{tab:params};][]{Clague.James.2002}.

@@

The modelled ice sheet interacts with the underlying bedrock in two ways.
Firstly, \emph{bedrock temperatures} are computed to a given depth by
appending the
ice enthalpy model (\ref{eqn:enthalpy}) with an underlying bedrock thermal
model. The only process accounted for is vertical heat conduction, hence this model
consists of a simple diffusion equation,
\begin{equation}
    \pdv{T}{t} = \frac{k_b}{\rho_b c_b} \Delta T \,,
\end{equation}
where $k_b$, $\rho_b$ and $c_b$ are the thermal conductivity, density and
specific heat capacity of the bedrock (\cref{tab:params}).
Bedrock temperature is conditioned at depth by a fixed upwards geothermal heat
flux $q_G$  (\cref{tab:params}). The bedrock thermal model
is necessary for modelling the Cordilleran ice sheet on multi-millennial
time-scales, because temperature fluctuations caused by climate change and
varying degrees of isolation from atmospheric temperatures by ice sheet
coverage, penetrate several kilometres
into the rock, and eventually affect basal ice temperatures.

Secondly, bedrock elevation evolves in response to the ice weight. The
\emph{bedrock deformation} model used in this thesis describes the flexure of
an elastic
lithosphere on top of an infinite half-space viscous mantle
\citep{Lingle.Clark.1985}. It can be described by a single pseudo-differential
equation of the bed elevation, $b$,
\begin{equation}
    \label{eqn:beddef}
    2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb} \,,
\end{equation}
where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity (\cref{tab:params}), and $\sigma_{zzb}$
corresponds to the ice load
\citep{Bueler.etal.2007}. Here, ${\Delta^2 = (\div\grad{})^2}$ designs the
biharmonic (Laplacian square) operator of
linear elastic theory, while $|\grad{}|$ is a pseudo-differential operator
defined through its Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. On
the left-hand side of \cref{eqn:beddef}, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response.

@@

In the simulations presented here, \cref{eqn:calovgreve} is numerically
approximated using week-long sub-intervals. For each sub-interval, PDDs, which
serve as a proxy for the available energy for melt, define the amount of snow
melt through a proportionality factor $F_s$. If all snow is melted, remaining
PDDs are then used to melt ice, using a proportionality factor $F_i$
(\cref{fig:model-pdd}). The degree-day factors for snow, $F_s$, and ice, $F_i$
(\cref{tab:params}),
are derived from mass-balance measurements on contemporary glaciers in the
Coast and Rocky mountains of British Columbia \citep{Shea.etal.2009}. In
{\CCLI}, the temperature standard deviation $\sPDD$ is a constant model
parameter. In {\CCYC}, it is computed from daily temperature values from the
North American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, using
a method initially developed in {\PSDV}, and further improved upon in {\PSDP}.
Although the PDD model is a crude parametrisation of processes happening at the
surface of a glacier, its applicability has been demonstrated
\citep[e.g.][]{Hock.2003},
while other, physically-based methods \citep[e.g.][]{Hock.2005}, require a
precise knowledge of the physical state of the atmosphere, which typically is highly
uncertain for past periods such as the last glacial cycle.

@@

@@

\begin{table*}
  \centering
  \caption{Sources of forcing data used to inform the ice sheet model.}
  \label{tab:datasources}
  \makebox[0pt]{\begin{tabular*}{170mm}{l@{\extracolsep{\fill}}lll}
    \tophline
    Name        & Type & Original reference & Employed in \\
    \middlehline

@@

    \bottomhline
  \end{tabular*}}
\end{table*}

The present-day monthly climatology is computed from
near-surface air temperature and precipitation rate fields extracted from
observational data and atmospheric reanalyses (\cref{tab:datasources}).
The lapse-rate correction, ${\Delta}T_{LR}$, is computed from the ice sheet
surface elevation, $s$, in relation to the
climate input reference bedrock topography, $b_{ref}$,
\begin{align}
    {\Delta}T_{LR}(t, x, y) = -\gamma\,[&s(t, x, y) - b_{ref}(x, y)] \\
                            = -\gamma\,[&h(t, x, y) + b(t, x, y) \notag \\
                                        &- b_{ref}(x, y)] \,,
\end{align}
thus accounting for the evolution of ice thickness, ${h=s-b}$, on the one hand,
and
for differences between the the ice flow model basal topography, $b$, and the
climate input topography, $b_{ref}$, on the other hand. All simulations use an
annual temperature lapse rate of $\gamma = 6\,\unit{K\,km^{-1}}$
(\cref{tab:params}).

@@

%\begin{table}
%  \centering
%  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
%           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
%           PROJ.4 parameter names, as used in all model runs, ice volume and
%           area calculations, and most figures.}
%  \label{tab:params}
%  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
%    \hline
%    Parameter               & Name              & Value \\
%    \hline
%    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
%    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
%    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
%    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
%    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
%    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
%    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
%    \hline
%  \end{tabular*}}
%\end{table}

@@

Next to the atmospheric forcing, another critical source of uncertainty in
palaeo-ice sheet modelling is the choice of parameters used by the surface mass
balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
are the degree-day factors for snow, $F_s$, and ice, $F_i$
(\cref{fig:model-pdd,tab:params}), and the standard deviation of
near-surface air temperature, $\sPDD$
(\cref{eqn:calovgreve,tab:params}). Although the first two can only be
deduced from surface mass-balance measurements on present-day glaciers
\citep[e.g.,][]{Shea.etal.2009} and assumptions that these values are also
valid for palaeo-simulations, $\sPDD$ can be estimated from temperature time
series \citep[e.g.,][]{Fausto.etal.2011}.
